---
title: "Lupine volatiles"
author: "Jacob Heiling, John Powers"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>
```
```{r setup, include=FALSE}
library(reshape2)
library(tidyverse)
#remotes::install_github("jmpowers/bouquet", build_vignettes = T)
library(bouquet)
library(vegan)
library(ggvegan)
library(pheatmap)
library(viridis)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning=F, message=F, 
                      fig.path = "plots/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=10, fig.width=10)
```

# Read scent data

```{r read_scents}
# 2017-2021 data

#reading in the Shimadzu search output is slow, skip it
# source("read_shimadzu.R")
# setwd("../Maxfield/data")
# datafiles <- list.files(pattern="Lupinus"); datafiles %>% set_names() %>% map(read.shimadzu) %>% bind_rows(.id="batch")
# lup.data <-  read.shimadzu("Lupinus_2017thru2021.txt") 
# setwd(path)
# save(lup.data, file="data/lupinus_gc_2017thru2021.rda")

load("data/lupinus_gc_2017thru2021.rda")
lup.data1721 <- lup.data; remove(lup.data)

# 2023 and 2025 (sampled 2024) data
# source("read_shimadzu.R")
# setwd("./data/2023_2025/")
# datafiles <- list.files(pattern=".txt")
# lup.data2325 <- datafiles %>% set_names() %>% map(read.shimadzu) %>% bind_rows()
# setwd("../..")
#save(lup.data2325, file="data/lupinus_gc_2023_2025.rda")

load("data/lupinus_gc_2023_2025.rda")
lup.data <- bind_rows(lup.data1721, lup.data2325)

# load short names and standard regressions
lupchems <- read_tsv("data/lupmeta2017thru2025 - lupchem.tsv") %>% 
  select(name, shortname, standard) %>% 
  mutate(class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
                            Sesquiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
  left_join(read_csv("data/regressions_181921_filtered_slopes.csv") %>% filter(year=="2019"))
class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(lupchems$class))

#shorten chemical names and merge compounds with multiple names
shortnames <- lupchems %>% select(name, shortname) %>% filter(shortname!="") %>% deframe()
#shortnames[shortnames %in% shortnames[duplicated(shortnames)]]
lup.data$Name <- recode(lup.data$Name, !!!shortnames)

lup.all <- lup.data %>% 
  #filter(str_detect(Filename, "Lupinus|Blank")) %>% 
  dcast(Filename~Name, sum, value.var="Area")
rownames(lup.all) <- lup.all[,1]
lup.all[,1] <- NULL
lup.cut <- lup.all[,colSums(lup.all)>1e8]#arbitrary cutoff 5e7
```

```{r kmeans}
k <- 7
set.seed(1)
km <- kmeans(decostand(lup.cut, method="log"), k, nstart=3)

lup.km <- tibble(FileName=rownames(lup.all)) %>% 
  mutate(rowSum = rowSums(lup.all),
         Type =  str_detect(FileName, "Blank") %>% if_else("blank","sample") %>% factor,
         #str_extract(FileName, "Blank|Lupinus") %>% fct_recode(blank="Blank",sample="Lupinus")
         nameBlank = Type=="blank",
         runYear = str_extract(FileName, "20[0-9][0-9]") %>%factor,#replace_na("2018") %>%
         Cluster = km$cluster) %>% # Figure out which k-means clusters are the blanks
  mutate(kBlank = Cluster %in% (count(., nameBlank, Cluster) %>% filter(nameBlank, n>2) %>% pull(Cluster)),
         Mixup = nameBlank != kBlank)

lup.km %>% count(kBlank, nameBlank) %>% kable(caption = "k-means attempt to separate blanks")
#Blanks are in two clusters near other samples, so kmeans not working very well to distinguish them (unless high n)
```

#  Read metadata

Samples dropped (blanks, leaks, nonfloral, or other users):

```{r metadata}
# Moved this section from markes_sequence.R, code above that section outputs this Rdata:
#load("data/markes_sequence.rda")#../Inventory/ #old format for skips
#now gcYEAR.csv files from the RMBL-GCMS project have the unannotated skip info 
sequ.summary <- list.files("./data/markes_sequence/", pattern=".csv", full.names=T) %>% read_csv()

lupinus <- sequ.summary$FileName %in% lup.km$FileName # #str_detect(sequ.summary$FullName, "Lupinus|lupinus|Lupine")
lupinus.batchids <- sequ.summary %>% filter(lupinus) %>% select(id) %>% unique() %>% na.omit() 
lupgc <- sequ.summary %>% 
  filter(id %in% lupinus.batchids$id | lupinus) %>% #get entire batch if it had a sample that matches 
  left_join(lup.km %>% select(FileName, rowSum, nameBlank, Mixup, kBlank, Cluster), by="FileName") %>% 
  write_csv("output/lupinus_all_2017thru2025.csv")

pops <- read_tsv("data/lupmeta2017thru2025 - pops.tsv") %>% mutate(population=factor(population, levels=unique(population)))
popnames <- pops %>% select(pop,population) %>% deframe()

# Get metadata from filenames in "RMBL GCMS Data Inventory"
lupmeta <- list.files("./data/markes_sequence/annotated", patter=".csv", full.names=T) %>% map_dfr(~read_csv(.x, na="NA")) %>% 
#read_tsv("data/RMBL GC-MS Data Inventory - maxfield_all210827annot.tsv", na="NA") %>% 
  filter(user=="J Heiling : Lupinus") %>% #TODO not going to work since user is based on filename, not sample
  distinct(FileName, .keep_all = TRUE) %>% 
  mutate(sample = na_if(sample,"") %>% coalesce(FileName) %>% str_remove(".qgd")) %>% 
  select(index, FileName, batch, sample) %>% 
  mutate(type = ifelse(str_detect(sample,"air"), "ambient", "floral"),
         sample_grid = str_remove(sample, "air_"), scented=NA_character_) %>% 
  separate(sample_grid, into=c("species","pop","sample_year","plant","rundate","GCn"), remove=F) %>% 
  full_join(lupgc %>% distinct(FileName, .keep_all = T) %>% select(-sample), by="FileName") %>% #was creating sample.x, sample.y so drop that column
  # lupmeta %>% select(index, FileName, sample, rowSum, Cluster, type, pop, sample_year, plant, rundate, GCn) %>% 
  #   drop_na(Cluster) %>% 
  #   separate(FileName, into=paste0("fn",1:8), remove=F, fill="right") %>% 
  #   write_csv("output/lupmeta.csv", na="") #fill in metadata for old samples (2017-2019)
  rows_update(read_tsv("data/lupmeta2017thru2025 - lupmeta.tsv", col_types="cccccccccccccccccccc") %>% 
                #filter(is.na(index)) %>% #just dropped this-restricts to 2017-2019
                select(FileName, type, pop, sample_year, plant, scented, rundate), by="FileName") %>% 
  mutate(index = coalesce(index, 10000 + 1:nrow(.)),
         sample = coalesce(sample, FileName) %>% str_remove(".qgd")) %>% #these filled samples were excluded above with na_if(sample,"")
  drop_na(nameBlank) %>% #drops rows in same batchids but not integrated (other people's samples)
  filter(!nameBlank) %>% #!kBlank
  filter(type %in% c("ambient","floral")) %>% #exclude leaf controls and insects
  left_join(pops) %>% mutate(pop = factor(pop, levels=pops$pop))

(dropped_samples <- sort(setdiff(levels(lup.data$Filename), lupmeta$FileName)))

with(lupmeta, table(population, sample_year))#sample_years are rough - split from filename
lupmeta %>% filter(type=="floral") %>% count(population, sample_year, scented) %>% pivot_wider(names_from=scented,values_from=n) %>% 
  kable(caption = "Inventory of scented (1) or unscented (0) samples")

lup <- lup.all[lupmeta$FileName,]
rownames(lup) <- rownames(lupmeta) <- lupmeta$index
save(lup, lupmeta, file="output/lupinus_volatiles_2017thru2025.rda")
```

# Compare all samples and blanks 

```{r blanks}
lup.nmds <- metaMDS(sqrt(lup.cut), dist="bray", autotransform = FALSE, trace=F)
lup.nmds.points <- fortify(lup.nmds) %>% as_tibble() %>% 
  filter(score=="sites") %>% left_join(lup.km, by=c("label"="FileName"))

nmds.plot <- ggplot(lup.nmds.points, aes(x=NMDS1, y=NMDS2, shape=Type)) + theme_classic()
nmds.plot + geom_point(aes(color=log(rowSum)), size=2) + scale_color_viridis_c()
nmds.plot + geom_point(aes(color=runYear), size=2)
nmds.plot + geom_point(aes(color=factor(Cluster)), size=2)
nmds.plot + geom_point(aes(color=nameBlank), size=2)

lup.cap <- capscale(lup.cut ~ nameBlank, distance="bray", metaMDSdist = F, data=lup.km)
anova(lup.cap)
lup.cap.points <-  fortify(lup.cap) %>% as_tibble() %>% 
  filter(score=="sites") %>% left_join(lup.km, by=c("label"="FileName"))
ggplot(lup.cap.points, aes(x=CAP1, y=MDS1, alpha=nameBlank)) + geom_point() +
  scale_alpha_manual(values=c(0.2,1))
cap.chems <- fortify(lup.cap) %>% as_tibble() %>% filter(score=="species") %>% arrange(desc(CAP1))
```

# Compare flowers and ambient controls

```{r ambients}
with(lupmeta, table(type, sample_year))

lup.nmds.noblanks <- metaMDS(lup[,colSums(lup)>1e8], distance="bray", autotransform=F, trace=F)
lup.nmds.points <-  as_tibble(fortify(lup.nmds.noblanks))
lup.nmds.points %>% filter(score=="sites")%>% bind_cols(lupmeta) %>% 
ggplot(aes(x=NMDS1, y=NMDS2)) + 
  geom_label(aes(label=sample, color=type)) +
  geom_text(data= lup.nmds.points %>% filter(score=="species"), aes(label=label))

lup.cap.noblanks <- capscale(lup[,colSums(lup)>1e8] ~ type, distance="bray", metaMDSdist = F, data=lupmeta)
anova(lup.cap.noblanks, by="term")
lup.cap.points <-  as_tibble(fortify(lup.cap.noblanks))
lup.cap.points %>% filter(score=="sites")%>% bind_cols(lupmeta) %>% 
  ggplot(aes(x=CAP1, y=MDS1)) + 
  geom_label(aes(label=sample, color=paste(type, sample_year))) +
  geom_text(data= lup.cap.points %>% filter(score=="species"), aes(label=label))
```

# Filtering

```{r filtering}
lup.data.cut <- lup.data[lup.data$Filename %in% lupmeta$FileName,]
#lup.data.cut$Filename <- lupmeta$sample[match(lup.data.cut$Filename, lupmeta$FileName)] #don't use sample!
lup.data.cut <- lup.data.cut[lup.data.cut$Name != "", ]#drop peaks that were not identified
lup.data.cut$Name <- droplevels(lup.data.cut$Name)
longdata <- load_longdata(lup.data.cut, sample="Filename", RT="Ret.Time", name="Name", area="Area", match = "SI", maxmatch=100)
metadata <- load_metadata(lupmeta %>% rename(realsample=sample), 
                          sample="FileName", type="type")#was sample="sample"
metadata$type <- factor(metadata$type)
vol.all <- make_sampletable(longdata, metadata)
chems <- make_chemtable(longdata, metadata)

### Flower filtering
chemsf <- chems %>%
  filter_RT(2, 17) %>% 
  filter_match(0.8) %>% 
  filter_freq(0.1, group = FALSE) %>% 
  filter_contaminant(cont.list = c("Acetic acid, trichloro-, nonyl ester",
                                   "2-Pentanol, 3-chloro-4-methyl-, (R*,R*)-(.+/-.)-",
                                   "Hexyl chloroformate", "Oxirane, decyl-", #below: alkanes with many RTs
                                   "4-methylpentanoic acid", "hexadecane", "1-nonoxytetradecane",
                                   "13-methyltetradecanal", "nonylcyclohexane", "3,8-dimethyldecane",
                                   "farnesane", "2,5-dimethylnonane", "3,3,5-trimethylheptane",
                                   "3,3,5-trimethylcyclohexan-1-ol")) %>% #quant and quant never both nonzero
  filter_area(min_maximum = 1e6) %>% 
  filter_ambient_ratio(vol.all, metadata, ratio = 4) %>%
  filter_ambient_ttest(vol.all, metadata, alpha = 0.05, adjust = "fdr") %>%
  combine_filters() 
chemsf$filter_final <- with(chemsf, (filter_RT == "OK" & filter_match=="OK" & filter_freq.floral == "OK" & 
                              filter_contaminant == "OK" & filter_area == "OK" & 
                                filter_ambient_ratio == "OK") | name=="indole")

#Prune based on bouquet filters
vol <- prune_sampletable(vol.all, chemsf, metadata)
exclude_low <- c("JH_Lup_arg_BVB_24-23redo_7102025_03.qgd", "JH_Lup_arg_BVB_24-30_792025_03.qgd") #<1 ng/hr volatiles
meta <- metadata[metadata$type == "floral",][!rownames(vol) %in% exclude_low,]
vol <- vol[!rownames(vol) %in% exclude_low,] 
chm <- chemsf[match(colnames(vol), chemsf$name),]

#Convert peak areas to nanograms
lupchems2 <- lupchems[match(colnames(vol), lupchems$shortname),]
vol <- sweep(vol, 2, lupchems2$area_per_ng, FUN = '/')
vol <- vol / (40/60 + 20/60) # divide by equilibration + pumping time in hours (40+20 min)
meta$Total <- rowSums(vol)

#bouquet diagnostics
plot_filters(chemsf, option="prop")
plot_filters(chemsf, option="rarity")
plot_filters(chemsf, option="ambient")
plot_filters(chemsf, option="volcano")
```

```{r diverse, eval=FALSE}
tointegrate <- colnames(vol)
best <- list()
i <- 1
vol.left <- vol[,tointegrate]^(1/4) # halfway between areas and presence-absence
while(ncol(vol.left)>0) {
  print(dim(vol.left))
  best[[i]] <- which.max(rowSums(vol.left))
  vol.left <- vol.left[ , !decostand(vol.left[best[[i]],], method="pa"), drop=FALSE]
  i <- i+1
}
(best.vol <- as.data.frame(t(vol[unlist(best),tointegrate])))
```

# Quant

```{r quant}
#Load quantitative integrations
source("read_shimadzu.R")
quant.full <- read.shimadzu.quant("./data/lupinus_2017_2025_quant35.txt") %>%
  mutate(Name = trimws(Name), Area=replace_na(Area, 0)) %>%
  distinct(Filename, Name, .keep_all=T)

quant.full %>% drop_na(Ret.Time) %>% mutate(Name = fct_reorder(Name, Ret.Time)) %>%
  ggplot(aes(x=Ret.Time, y=Name)) + geom_jitter(width=0, size=0.2)
quant.full %>% drop_na(Ret.Time) %>% group_by(Name) %>% mutate(RT.median = median(Ret.Time)) %>%
  ggplot(aes(x=Ret.Time-RT.median, y=Name, size=Area)) + geom_jitter(width=0)

quant <- quant.full %>% select(Dirname, Filename, Name, Area) %>%
  mutate(Dirname = fct_relabel(Dirname, str_remove, "C:/GCMSsolution/Data/")) %>%
  mutate(Name = recode(Name, !!!shortnames)) %>%
  filter(Name != "3,3,5-trimethylcyclohexan-1-ol") %>% #no overlap between nonzero vol and quant
  pivot_wider(names_from = "Name", values_from="Area") %>% as.data.frame()

rownames(quant) <- quant$Filename
quant[,c("Dirname","Filename")] <- NULL
quant.all <- quant[rownames(vol),]
write_tsv(quant.all %>% rownames_to_column("Filename"), "output/lupinus_quant_all.tsv")

#compare areas of quant ions and qual peaks to find conversion factor
quant.slopes <- set_names(rep(0,ncol(quant.all)), colnames(quant.all))
for(compvol in colnames(quant.all)) {
  nozeros <- bind_cols(select(vol,qual=compvol), select(quant.all,quant=compvol)) %>% 
    filter(!(qual==0 | quant == 0))
  quant.slopes[compvol] = coef(lm(quant ~ 0 + qual, data=nozeros))["qual"]
}
write_tsv(enframe(quant.slopes) %>% left_join(enframe(shortnames, name="Name", value="name")),
          "output/quant_slopes.tsv")

quant.scaled <- sweep(quant.all, 2, quant.slopes, FUN = '/')

quant.scaled.vol <- quant.scaled[meta$sample, ]#sample is formerly FileName but renamed by bouquet

tibble(quant=unlist(vol[,colnames(quant.scaled.vol)]), qual=unlist(quant.scaled.vol)) %>% 
  ggplot(aes(x=qual, y=quant)) + geom_point(alpha=0.5) + geom_abline(color="red")+ 
  scale_x_sqrt("Qualitative (ng/hr)") + scale_y_sqrt("Quantitative (ng/hr)")

#swap in scaled quant integrations
#make sure not to run above code after this line has run, or start vol over with pruning
vol[,colnames(quant.scaled.vol)] <- quant.scaled.vol

save(vol, meta, chm, file="output/lupinus_filtered_2017thru2025.rda")
write_csv(vol, "output/lupinus_vol.csv")
write_csv(meta, "output/lupinus_meta.csv")
write_csv(chm, "output/lupinus_chm.csv")
```

# Retention times

```{r retention}
longdata.RT <- longdata %>% filter(name %in% chemsf$name[chemsf$filter_final]) %>% 
  left_join(select(metadata, sample, sample_year)) %>%
  group_by(name) %>%  mutate(n=n(), RT.sd=sd(RT)) %>% ungroup() %>% 
  mutate(name=fct_reorder(name, RT))

RT.year <- longdata.RT %>% group_by(name, sample_year) %>% summarize(RT=median(RT)) %>% pivot_wider(names_from=sample_year, values_from=RT) 
RT.year %>% filter(name %in% setdiff(colnames(vol), colnames(quant.all))) %>% arrange(`17`)

RT.tol <- 0.1  # set in quantitative params
ggplot(longdata.RT, aes(x=RT, y=name,color=RT.sd)) + geom_point(position=position_jitter(width=0, height=0.5)) + 
  scale_color_viridis_c(option="magma") + theme_dark()+
  geom_text(aes(x=4.5, label=n), color="grey80") +
  stat_summary(fun="median", fun.max=~median(.x)+RT.tol, fun.min=~median(.x)-RT.tol, geom="pointrange", color="white", size=0.5, shape=15)

ggplot(longdata.RT,  aes(y=RT, x=factor(sample_year), color=RT.sd)) + 
  facet_wrap(vars(name), scales="free_y") + geom_boxplot() + scale_color_viridis_c(option="magma") + theme_dark() 
```

```{r mess, eval=FALSE}
mess <- c("allo-ocimene", "carveol", "cosmene", "3,4-dimethylocta-2,4,6-triene")
longdata.RT.mess <- longdata.RT %>% filter(name %in% mess) %>%
  arrange(RT) %>% mutate(cluster = kmeans(RT, 3, nstart=4)$cluster) %>%
  group_by(cluster) %>% mutate(RT.cluster = round(mean(RT),2)) %>% ungroup()

#shows 4 peaks that come up & their retention times. Color shows you what compound it gets identified as
ggplot(longdata.RT.mess, aes(x=factor(RT.cluster), fill=name)) + geom_bar()
#are there differences in retention times? answer: no
ggplot(longdata.RT.mess, aes(x=RT, color=name, fill=name)) + facet_wrap(vars(RT.cluster), scales="free_x") + theme_minimal() +
  geom_density(alpha=0.2, linewidth=2) + geom_jitter(aes(y=-as.integer(factor(name))*10), width=0.0002, height=4)

mess.wide <- longdata.RT.mess %>% select(sample, RT.cluster, area) %>% pivot_wider(names_from = RT.cluster, values_from=area, values_fn=sum, values_fill=0)
```


# Heatmap

```{r heatmap,  dev="png", dev.args=list(), fig.width=16}
pheatmap(t(vol^(1/5)), color = mako(512),
         annotation_col = select(meta, sample_year, population, sample) %>% 
           column_to_rownames("sample") %>% mutate(sample_year=factor(sample_year)),
         annotation_row = select(lupchems, class, shortname) %>% distinct(shortname, .keep_all=TRUE) %>% column_to_rownames("shortname"))
```

# NMDS of filtered volatiles

```{r nmds_filtered}
lup.nmds <- metaMDS(sqrt(vol), dist="bray", autotransform = FALSE, trace=F)
lup.nmds.points <-  fortify(lup.nmds) %>% as_tibble()
lup.nmds.samples <- lup.nmds.points %>% filter(score=="sites") %>% bind_cols(meta)
lup.nmds.chems <- lup.nmds.points %>% filter(score=="species")

nmds.plot <- ggplot(lup.nmds.samples, aes(x=NMDS1, y=NMDS2)) + 
  geom_text(data=lup.nmds.chems, aes(label=label)) + theme_classic() #alpha=chm$freq.floral

nmds.plot + geom_label(aes(label=paste0(pop, plant), color=pop)) + guides(color="none")
nmds.plot + geom_point(aes(color=sqrt(rowSums(vol)), shape=sample_year), size=2) + scale_color_viridis_c("Sqrt total peak area", option="magma") 
nmds.plot + geom_point(aes(color=sample_year), size=2)
```

# Total volatiles by run date

```{r rundate}
ggplot(meta, aes(x=paste0("20", sample_year, " run ", rundate), color=sample_year, y=Total)) + 
  geom_violin(scale="width", draw_quantiles=(1:3)/4) + geom_point() +
  theme(axis.text.x = element_text(angle=90))
```

# Correlated volatiles

```{r corr}
ggplot(cbind(vol,meta), aes(x=`(E)-b-ocimene`, y=`(Z)-b-ocimene`, color=sample_year)) + geom_point() + geom_smooth(se=F) + scale_y_sqrt() + scale_x_sqrt()
ggplot(cbind(vol,meta), aes(x=`(E)-b-ocimene`, y=`b-myrcene`, color=sample_year)) + geom_point() + geom_smooth(se=F) + scale_y_sqrt() + scale_x_sqrt()
```

# CAP by GC run year

Only look at 2019 vs. 2020 here
```{r cap_year}
twoyears <- meta$sample_year %in% c("19","20")
lup.cap <- capscale(sqrt(vol[twoyears,]) ~ sample_year, dist="bray", data=meta[twoyears,])
#View(lup.cap$CCA$v)
lup.cap.points <-  fortify(lup.cap) %>% as_tibble()
lup.cap.samples <- lup.cap.points %>% filter(score=="sites") %>% bind_cols(meta[twoyears,])
lup.cap.chems <- lup.cap.points %>% filter(score=="species")
ggplot(lup.cap.samples, aes(x=CAP1, y=MDS1)) + 
  geom_point(aes(color=sample_year)) + 
  geom_text(data=lup.cap.chems, aes(label=label, alpha=chm$freq.floral)) + theme_classic()
```

# PERMANOVA of populations

```{r permanova_pop}
adonis2(sqrt(vol) ~ sample_year + population, data=meta)
```

# CAP of populations

```{r cap_pop}
lup.cap <- capscale(sqrt(vol) ~ sample_year + population, dist="bray", data=meta)
lup.cap
anova(lup.cap, by="margin")
lup.cap.points <-  fortify(lup.cap) %>% as_tibble()
lup.cap.samples <- lup.cap.points %>% filter(score=="sites") %>% bind_cols(meta)
lup.cap.chems <- lup.cap.points %>% filter(score=="species")

ggplot(lup.cap.samples, aes(x=CAP1, y=CAP2)) + 
  geom_text(data=lup.cap.chems, aes(label=label, alpha=chm$freq.floral)) +
  geom_label(aes(label=paste0(pop, plant), color=sample_year)) + theme_classic()

ggplot(lup.cap.samples, aes(x=CAP1, y=CAP2)) + 
  geom_point(aes(color=rowSum)) + scale_color_viridis_c("Total peak area") + theme_classic()
```

# Boxplot of populations

Compounds are in order of mean peak area, arranged in columns.

```{r boxplot, fig.width=16}
topamounts <- sort(colSums(vol), decreasing = T)
topchems <- head(names(topamounts), 18)

ggplot(enframe(topamounts) %>% mutate(name=fct_reorder(name, value)), aes(y=name, x=value)) + geom_col()

vol %>% select(topchems) %>% rownames_to_column("sample") %>% 
  pivot_longer(topchems) %>% mutate(name=fct_relevel(name, topchems)) %>% 
  left_join(meta) %>% 
  ggplot(aes(x=population, y=value, color=sample_year)) + facet_wrap(vars(name), ncol=3, scales="free_y", dir="v")+
  scale_y_sqrt() + scale_x_discrete(guide=guide_axis(angle=90))+
  geom_boxplot(position=position_dodge(width=0.9), outlier.shape = NA) + 
  geom_point(position=position_dodge(width=0.9)) +
  labs(x="Population", y="Peak area", color="Year")
```


# CAP of un/scented
```{r cap_scented}
vol.sniff <- vol[!is.na(meta$scented),]
meta.sniff <- meta[!is.na(meta$scented),]
lup.cap <- capscale(sqrt(vol.sniff) ~ scented, dist="bray", data=meta.sniff)
lup.cap
anova(lup.cap, by="margin")
lup.cap.points <-  fortify(lup.cap) %>% as_tibble()
lup.cap.samples <- lup.cap.points %>% filter(score=="sites") %>% bind_cols(meta.sniff)
lup.cap.chems <- lup.cap.points %>% filter(score=="species")

ggplot(lup.cap.samples, aes(x=CAP1, y=MDS1)) + 
  geom_text(aes(label=paste0(pop, plant), color=scented)) +
  geom_text(data=lup.cap.chems, aes(label=label, alpha=chm$freq.floral)) + theme_classic()

ggplot(lup.cap.samples, aes(x=CAP1, y=MDS1)) + 
  geom_point(aes(color=rowSum, shape=scented), size=3) +
  geom_text(data=lup.cap.chems, aes(label=label, alpha=chm$freq.floral))+ scale_color_viridis_c("Total peak area") + theme_classic()

ggplot(cbind(meta.sniff, vol.sniff), aes(x=pop, y=`(E)-b-ocimene`, color=scented)) + 
  geom_boxplot(position=position_dodge(width=0.9))+
  geom_point(position=position_dodge(width=0.9), size=3) +  scale_y_sqrt() + theme_minimal()

ggplot(cbind(meta.sniff, vol.sniff), aes(x=pop, y=rowSums(vol.sniff), color=scented)) +   geom_boxplot(position=position_dodge(width=0.9))+
  geom_point(position=position_dodge(width=0.9), size=3)+  scale_y_sqrt("Total emissions (ng/hr)") + theme_minimal() 
```

# CAP of elevation
```{r cap_elevation}
lup.cap <- capscale(sqrt(vol) ~ elev, dist="bray", data=meta)
lup.cap
anova(lup.cap, by="margin")
lup.cap.points <-  fortify(lup.cap) %>% as_tibble()
lup.cap.samples <- lup.cap.points %>% filter(score=="sites") %>% bind_cols(meta)
lup.cap.chems <- lup.cap.points %>% filter(score=="species")

ggplot(lup.cap.samples, aes(x=CAP1, y=MDS1)) + 
  geom_text(aes(label=paste0(pop, plant), color=elev)) + scale_color_viridis_c()+ 
  geom_text(data=lup.cap.chems, aes(label=label, alpha=chm$freq.floral)) + theme_classic()

ggplot(cbind(meta, vol), aes(x=elev, y=`(E)-b-ocimene`, color=sample_year, shape=scented, linetype=scented)) + geom_point() + 
  stat_summary(fun=mean, geom="line")+ stat_summary(fun=mean, geom="point", size=4)+
  scale_shape_manual(values=c(4,15), na.value=1) + scale_linetype_manual(values=c(2,1), na.value=1) +  scale_y_sqrt() + theme_minimal()+
  geom_text(data=distinct(pops,population,.keep_all = T), aes(x=elev,y=0,label=pop), nudge_y = -2, angle=90, inherit.aes = F)

ggplot(cbind(meta, vol), aes(x=elev, y=Total, color=sample_year, shape=scented, linetype=scented)) + geom_point() + 
  stat_summary(fun=mean, geom="line")+ stat_summary(fun=mean, geom="point", size=4)+ 
  scale_shape_manual(values=c(4,15), na.value=1) + scale_linetype_manual(values=c(2,1), na.value=1) +  scale_y_sqrt("Total emissions (ng/hr)") + theme_minimal()+
  geom_text(data=distinct(pops,population,.keep_all = T), aes(x=elev,y=0,label=pop), nudge_y = -2, angle=90, inherit.aes = F)
```